Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-16 BN_discretizer_paper.tex
misc

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

@@

Algorithm \ref{alg:disc_one} has an exponential $\bigo\paren{\paren{v'_\text{max}}^{n_p} \cdot N^2}$ runtime.
This exponential runtime is severely limiting for networks with a large number of discretization edges.
This section introduces an approximation to the objective function that significantly reduces the runtime and still preserves the quality of discretization.
The approximation replaces the dominator of the last factor in Equation \ref{eq:likelihood_one} with

@@

In Equation \ref{eq:example_approx}, each child carries two terms, as shown in the first set of square brackets, but the two parent variables only carry a total of three terms, as shown in the second set of square brackets.
In this case the child variables have greater effect than the parent variables even though the number of child and parent variables are the same.
However, if $C_2$ has an additional parent $S_2$, as shown in the right graph of Figure \ref{fig:example_networks}, the importances of $C_2$ will be reduced, because the information from $C_2$ is now affected by $S_2$.
This argument shows that the proposed method, both before and after approximation, indeed incorporates graph structure when discretizing continuous variables.

\begin{figure}[ht]
  \label{fig:example_networks}
  \begin{tabular}{cc}
    \input{graph2_left}
    \end{tabular}
   \hspace{5em}
    \begin{tabular}{cc}
    \input{graph2_right}
  \end{tabular}
  \caption{Two example networks}
\end{figure}

@@

\begin{equation}
M_X = \curly{X_\text{min}, X_\text{min} + \delta, X_\text{min} + 2\delta, \ldots, \delta, X_\text{max}}\text{,}
\end{equation}

@@

The section describes experiments designed to test the soundness of the proposed method. The experiments were run on four datasets from the Irvine repository \citep{Lichman:2013}. For each dataset, the experiment consists of two parts. In the first part, the Bayesian network structure is known in advance, which is obtained by equal-width prediscretization and the K2 algorithm. In the second part, the Bayesian netowrk structure is not provided in advance, and it is learned along with the discretization process.

For the case that network structure is given, a comparison between the proposed method and the MDL principle discretization is made. 10-fold cross validation of log-likelihood on data set $D_{\text{all}}$ is calculated as follows: firstly, learn discretization policies of each continuous variables from train data $D_{\text{train}}$. Secondly, obtain conditional probability tables $\phi_{x^* \mid \Pi^*_x}$ from the discretized train data $D^*_{\text{train}}$. Thirdly, For a test data instance $(x_1,x_2, \ldots, x_n)$ from $D_{\text{test}}$, the likelihood is
\begin{equation}
P(x_1, \ldots, x_n) = \prod_{i=1}^{n} \phi_{x^*_i \mid \Pi^*_{x_i}} \cdot \left( {1\over{M'_{X_i}[j_i+1] - M'_{X_i}[j_i]}} \right)^{\boldsymbol{1}_{ \left\lbrace X_i \text{ is continuous}  \right\rbrace }}
\end{equation}
, where $M'_{X_i}$ is the augmented discretization policy on continuous variable $X_i$ such that $M'_{X_i}[1] = min(D_{\text{all},X})$, $M'_{X_i}[j] = M_{X_i}[j-1]$ for $j=2,...,k_i$, and $M'_{X_i}[k_i+1] = max(D_{\text{all},X})$, $k$ is the number of intervals of continuous variable $X_i$ after discretization. Furthermore, $M'_{X_i}[j_i] \leq x_i < M'_{X_i}[j_i +1]$. This can be interpreted as follows: once a discrete-valued Bayesian network is learned from train data, test data are sampled from the learned network. If data is continuous, then it is uniformly sampled from the corresponding discretization interval.

\subsection{Dataset 1: Auto MPG} 
\label{subsec:autompg}

In this subsection, two experiments are conducted on the Auto MPG dataset from the Irvine repository \citep{Lichman:2013}. This dataset has 392 instances (delete the 6 instance with missing data) and eight variables. Three variables are discrete: $B$, $G$, and $H$ (in the following experiment, variables are named and ordered according to the dataset information webpage). They have 5, 13, and 3 cardinality numbers, respectively. In the first experiment, the network structure is given. In the second experiment, the network structure is not provided.

\subsubsection{Experiment 1 on Auto MPG dataset}
\label{subsubsec:autompg_exp1}
The proposed discretization method and the MDL discretization method are tested on the Bayesian network structure in Fig. \ref{Auto_graph_1}, which is obtained by running K2 algorithm 1000 times on prediscretized data, each time with a random restart. The prediscretization is done by equal-width discretization on continuous variables by 5 intervals (median of cardinalities of discrete variables). Shaded nodes in Fig. \ref{Auto_graph_1} means that they are discrete-valued variables.

\begin{figure}[ht]
\centering
    \begin{tabular}{cc}
      \input{graph_autocar_1}
    \end{tabular}
  \caption{The Bayesian network obatined by K2 algorithm and prediscretization on Auto MPG dataset}
  \label{Auto_graph_1}
\end{figure}

@@

\begin{table}
\centering
\scalebox{1.0}{
\input{table_autoMPG_1.tex}
}
\caption{Discretization result of Auto MPG dataset based on the graph in Fig. \ref{Auto_graph_1}}
\label{Auto_MPG_disc_table_1}
\end{table}

@@

\subsubsection{Experiment 2 on Auto MPG dataset}
\label{subsubsec:autompg_exp2}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

\newcommand{\discset}{\mathcal{M}}

@@

Algorithm \ref{alg:disc_one} has an exponential $\bigo\paren{\paren{v'_\text{max}}^{n_p} \cdot N^2}$ runtime which severely limits discretization in networks with a large number of discretization intervals or parent variables.
This section introduces an approximation to the objective function that significantly reduces the runtime and still preserves the quality of discretization.
The approximation replaces the dominator of the last factor in Equation \ref{eq:likelihood_one} with

@@

In Equation~\ref{eq:example_approx}, each child carries two terms, as shown in the first set of square brackets, but the two parent variables only carry a total of three terms, as shown in the second set of square brackets.
In this case the child variables have greater effect than the parent variables even though the number of child and parent variables are the same.
However, if $C_2$ has an additional parent $S_2$, as shown in the right graph of Figure~\ref{fig:example_networks}, the importances of $C_2$ will be reduced, because the information from $C_2$ is now affected by $S_2$.
This argument shows that the proposed method, both before and after approximation, indeed incorporates graph structure when discretizing continuous variables.

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph2_left}
    \end{tabular}
   \hspace{5em}
    \begin{tabular}{cc}
    \input{graph2_right}
  \end{tabular}
  \caption{Two example networks}
  \label{fig:example_networks}
\end{figure}

@@

\begin{equation}
M_X = \curly{X_\text{min}, X_\text{min} + \delta, X_\text{min} + 2\delta, \ldots, X_\text{max}}\text{,}
\end{equation}

@@

This section describes experiments conducted to evaluate the proposed methods.
All experiments were run on datasets from the publically available Irvine machine learning repository \citep{Lichman:2013}.
Variables are labelled alphabetically in the order given on the dataset information webpage.
Subscripts inducate the number of discretized intervals.
Shaded nodes are originally discrete and unshaded nodes are originally continuous.

For each dataset, the experiment consists of two parts.
In the first part, the Bayesian network structure is known in advance, which is obtained by equal-width prediscretization and the K2 algorithm.
In the second part, the Bayesian netowrk structure is not provided in advance, and it is learned along with the discretization process.
\tim{We may not want to state this up front so we can modify what we show for each dataset.}

The discretizations are compared using the mean cross validated log-likelihood of the data set $D$ given the resulting graph structure $G$ and discretization policy set $\discset$.
The log-likelihood contains two components,

\begin{equation}
\ln p(D\mid G, \discset) = \ln P(D^*\mid G) + \ln p(D\mid \discset, D^*)\text{,}
\end{equation}

\noindent
where $D$ is the original dataset and $D^*$ is the dataset discretized according to $\discset$.
The log-likelihood of the discretized dataset is the Bayesian score evaluated with a uniform prior~\citep{neapolitan2004learning}:

\begin{equation}
  \ln P(D^*\mid G) = \sum_{i=1}^n \sum_{j=1}^{q_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ij0}}
    }{
      \Gamma\paren{\alpha_{ij0} + m_{ij0}}
    }
  } + \sum_{k=1}^{r_i} \ln\paren{
    \frac{
      \Gamma\paren{\alpha_{ijk} + m_{ijk}}
    }{
      \Gamma\paren{\alpha_{ijk}}
    }
  }\text{.}
\end{equation}

\tim{We need to make sure everything is defined.}

The log-likelihood of the original dataset given the discrete dataset is

\begin{equation}
  \ln p(D\mid \discset, D^*) = \sum_{i=1}^n \sum_{j=1}^{q_i} \sum_{k=1}^{r_i} m_{ijk} \ln\paren{
    \frac{
      1
    }{
      \discset^{(ij)}_{k+1} - \discset^{(ij)}_{k}
    }
  }\text{.}
\end{equation}

\subsection{Auto MPG Dataset}

In this subsection, two experiments are conducted on the Auto MPG dataset from the Irvine repository \citep{Lichman:2013}.
The dataset has \num{392} instances after deleting the six instances with missing data, and eight variables.
Three variables are discrete: $B$, $G$, and $H$ with cardinality 5, 13, and 3 respectively.
Two expiriments are conducted, one with a given network structure and one in which the structure is learned.

\subsubsection{Structure is Given}

The proposed discretization method and the MDL discretization method are tested on Auto MPG data given the network network in Fig. \ref{Auto_graph_1}.
The structure was obtained by prediscretizing each continuous variable into five uniform-width intervals, where five is the median cardinality of the discrete variables, and then taking the most frequent structure from a thousand runs of K2.

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph_autocar_1}
  \end{tabular}
  \caption{The Bayesian network obtained with K2 on the prediscretized Auto MPG dataset}
  \label{Auto_graph_1}
\end{figure}

@@

\begin{table}[h]
  \centering
  \input{table_autoMPG_1.tex}
  \caption{Discretization result of Auto MPG dataset based on the graph in Fig. \ref{Auto_graph_1}. The discretization policy for the heuristic matches that for the optimal Bayesian approach. \todo{Need to explain why MDL fails}}
  \label{Auto_MPG_disc_table_1}
\end{table}

@@

\subsubsection{Structure is Learned}
